---
title: "CDC Data Exercise"
---

For this exercise, the dataset I'll be working on is called the [NCHS - Leading causes of Death: United States](https://data.cdc.gov/National-Center-for-Health-Statistics/NCHS-Leading-Causes-of-Death-United-States/bi63-dtpu/about_data). In summary, this dataset presents the age-adjust death rates for the 10 leading causes of death in the United States since 1999. 

Published by the **National Center for Health Statistics**, the dataset is described as follows:

*"Data are based on information from all resident death certificates filed in the 50 states and the District of Columbia using demographic and medical characteristics. Age-adjusted death rates (per 100,000 population) are based on the 2000 U.S. standard population. Populations used for computing death rates after 2010 are postcensal estimates based on the 2010 census, estimated as of July 1, 2010. Rates for census years are based on populations enumerated in the corresponding censuses. Rates for non-census years before 2010 are revised using updated intercensal population estimates and may differ from rates previously published.*

*Causes of death classified by the International Classification of Diseases, Tenth Revision (ICD–10) are ranked according to the number of deaths assigned to rankable causes. Cause of death statistics are based on the underlying cause of death."*

**Load Libraries**
```{r setup, warning = FALSE, message = FALSE} 
library(tidyverse)
library(dplyr)
library(ggplot2)
library(here)
library(gridExtra)
```

**Load Data**
```{r}
raw_data <- read.csv2(here("cdcdata-exercise", "NCHS_-_Leading_Causes_of_Death__United_States_20260209.csv"), sep = ",")
```

**Check Data**
```{r}
nrow(raw_data) #counting rows = counting entries
ncol(raw_data) #counting columns = counting variables
str(raw_data) #check structure
head(raw_data) #check top of data
tail(raw_data) #check bottom of data
colSums(is.na(raw_data)) #check for missing data
unique(raw_data$Year)
unique(raw_data$State) 
unique(raw_data$Cause.Name)
unique(raw_data$X113.Cause.Name)
ggplot(raw_data, aes(x = State)) +
    geom_bar()+
    theme(axis.text.x = element_text(angle = 90))
```

From checking the data, here's what stood out to me:

_1. Variables "Deaths" and "Age.adjusted.Death.Rate" are numbers but are stored as characters (I assume) because of commas._

_2. Fortunately, the dataset doesn't appear to have missing data._

_3. Year only runs from 1999 to 2017._

_4. 52 states are listed because in addition to the 50 states, the variable State also contains "District of Columbia" and "United States. At the very least, the dataset is consistent that the entries for each state is equal based on the figure."_

_5. It seems like there are two variables that describe the same thing: cause of death. It's just that one is more simplified than the other._

**Clean Data**
```{r}
#Duplicating dataframe to clean up
clean_data <- raw_data

#converting variables "Deaths" and "Age.adjusted.Death.Rate" into integer and numerical variables
#Using gsub to remove the commas
clean_data$Deaths <- as.integer(gsub(",", "", clean_data$Deaths))
clean_data$Age.adjusted.Death.Rate <- as.numeric(gsub(",", "", clean_data$Age.adjusted.Death.Rate))
```

**Exploratory Data Analysis**

```{r}
#Let's see how each cause of death did throughout the years

clean_data %>% 
    group_by(Cause.Name) %>%
    filter(State == "United States") %>%
    summarise(total_deaths = sum(Deaths)) %>%
    arrange(desc(total_deaths))

P1 <- clean_data %>%
    filter(State == "United States") %>%
    filter(Cause.Name == "Heart disease") %>%
    ggplot(aes(x = Year, y = Deaths)) +
    geom_line()+
    theme_classic()+
    labs(title = "Heart Disease")

P2 <- clean_data %>%
    filter(State == "United States") %>%
    filter(Cause.Name == "Cancer") %>%
    ggplot(aes(x = Year, y = Deaths)) +
    geom_line()+
    theme_classic()+
    labs(title = "Cancer")

P3 <- clean_data %>%
    filter(State == "United States") %>%
    filter(Cause.Name == "Stroke") %>%
    ggplot(aes(x = Year, y = Deaths)) +
    geom_line()+
    theme_classic()+
    labs(title = "Stroke")

P4 <- clean_data %>%
    filter(State == "United States") %>%
    filter(Cause.Name == "CLRD") %>%
    ggplot(aes(x = Year, y = Deaths)) +
    geom_line()+
    theme_classic()+
    labs(title = "CLRD")

P5 <- clean_data %>%
    filter(State == "United States") %>%
    filter(Cause.Name == "Unintentional injuries") %>%
    ggplot(aes(x = Year, y = Deaths)) +
    geom_line()+
    theme_classic()+
    labs(title = "Unintentional Injuries")

P6 <- clean_data %>%
    filter(State == "United States") %>%
    filter(Cause.Name == "Alzheimer's disease") %>%
    ggplot(aes(x = Year, y = Deaths)) +
    geom_line()+
    theme_classic()+
    labs(title = "Alzheimer's Disease")

P7 <- clean_data %>%
    filter(State == "United States") %>%
    filter(Cause.Name == "Diabetes") %>%
    ggplot(aes(x = Year, y = Deaths)) +
    geom_line()+
    theme_classic()+
    labs(title = "Diabetes")
    
P8 <- clean_data %>%
    filter(State == "United States") %>%
    filter(Cause.Name == "Influenza and pneumonia") %>%
    ggplot(aes(x = Year, y = Deaths)) +
    geom_line()+
    theme_classic()+
    labs(title = "Influenza and Pneumonia")

P9 <- clean_data %>%
    filter(State == "United States") %>%
    filter(Cause.Name == "Kidney disease") %>%
    ggplot(aes(x = Year, y = Deaths)) +
    geom_line()+
    theme_classic()+
    labs(title = "Kidney Disease")

P10 <- clean_data %>%
    filter(State == "United States") %>%
    filter(Cause.Name == "Suicide") %>%
    ggplot(aes(x = Year, y = Deaths)) +
    geom_line()+
    theme_classic()+
    labs(title = "Suicide")

grid.arrange(P1, P2, P3, P4, P5, P6, P7, P8, P9, P10, ncol = 2, nrow = 5)

#Let's see how they all did when graphed together.

clean_data %>%
    filter(State == "United States") %>%
    filter(!Cause.Name == "All causes") %>%
    ggplot(aes(x = Year, y = Deaths, group = Cause.Name, color = Cause.Name)) +
    geom_line()+
    theme_classic()

#Being in Georgia, let's see the top causes of death in Georgia and see if the order would match the order seen when looking across the whole country.

clean_data %>%
    group_by(Cause.Name) %>%
    filter(State == "Georgia") %>%
    summarise(total_deaths = sum(Deaths)) %>%
    arrange(desc(total_deaths))

```

From our exploratory data analysis, these are what stood out to me:

1. A couple of causes of death, like heart disease and stroke, actually dipped toward 2010 and slowly climbed back up.

2. On the other hand, few other causes of death steadily increased across time (e.g., cancer, unintentional injuries, CLRD, Alzheimer's, suicide.

3. Looking at them all together, we get a better sense of how vast the difference is in the deaths caused by heart disease and cancer compared to all other causes of death.

3. When looking at Georgia compared to the national perspective:
    a. unintentional injuries were more prevalent than CLRD,
    b. kidney diseases were more prevalent than influenza and pneumonia.

**Specific Questions**
_1. Which states have the highest and lowest death rates from cancer?_
```{r}
clean_data %>%
    filter(Cause.Name == "Cancer") %>%
    ggplot(aes(x = reorder(State, Age.adjusted.Death.Rate), y = Age.adjusted.Death.Rate)) +
    geom_boxplot()+
    theme_classic()+
    theme(axis.text.x = element_text(angle = 90))

clean_data %>%
    filter(!State == "United States") %>%
    filter(Cause.Name == "Cancer") %>%
    ggplot(aes(x = reorder(State, Deaths), y = Deaths)) +
    geom_boxplot()+
    theme_classic()+
    theme(axis.text.x = element_text(angle = 90))
```

It's interesting to see conclusions based on looking at more than one variable. For example, just by looking at the number of deaths due to cancer, we see that California has the highest number of deaths due to cancer than other states. However, if we look at the death rates, California actually ranks really low. In fact, the top three states with the most deaths due to cancer (California, Florida, Texas), are actually on the bottom half when looking at death rates. In fact, let's take a look at numbers for death rates in cancer across states.

```{r}
#mean deaths and death rates by State
clean_data %>%
  group_by(State) %>%
  summarise(
    avg_death_rate = mean(Age.adjusted.Death.Rate),
    avg_deaths     = mean(Deaths)) %>%
        print(n=52)
    
#mean deaths and death rates by Year
clean_data %>%
  group_by(Year) %>%
  summarise(
    avg_death_rate = mean(Age.adjusted.Death.Rate),
    avg_deaths     = mean(Deaths))
```

_2. Cancer has gone up over the years. How has the death rate been?_
```{r}
P2b <- clean_data %>%
    filter(State == "United States") %>%
    filter(Cause.Name == "Cancer") %>%
    ggplot(aes(x = Year, y = Age.adjusted.Death.Rate)) +
    geom_line()+
    theme_classic()

grid.arrange(P2, P2b, ncol = 2, nrow = 1)
```

So despite the number of deaths due to cancer going up, we've actually had amazing advances to reduce cancer death rates.

**This part was contributed by Aylla von Ermland:**

The AI prompt used to generate the synthetic data was: 

"*Write clean, reproducible R code (tidyverse, no loops) that generates a synthetic U.S. mortality dataset based on U.S. resident death certificates using the mean distribution of a real dataset provided below. Dataset structure is columns: Year, 113 Cause Name, Cause Name, State, Deaths, Age-adjusted Death Rate.*"

"*Years are between 1999–2017 (19 years total) and states are 52 in total (Alabama, Alaska, Arizona, Arkansas, California, Colorado, Connecticut, Delaware, District of Columbia, Florida, Georgia, Hawaii, Idaho, Illinois, Indiana, Iowa, Kansas, Kentucky, Louisiana, Maine, Maryland, Massachusetts, Michigan, Minnesota, Mississippi, Missouri, Montana, Nebraska, Nevada, New Hampshire, New Jersey, New Mexico, New York, North Carolina, North Dakota, Ohio, Oklahoma, Oregon, Pennsylvania, Rhode Island, South Carolina, South Dakota, Tennessee, Texas, United States, Utah, Vermont, Virginia, Washington, West Virginia, Wisconsin, Wyoming.)*"

"*Causes of death are 11 total, include both simplified and detailed ICD-style names: All causes, Heart disease, Cancer, Stroke, CLRD, Unintentional injuries, Alzheimer’s disease, Diabetes, Influenza and pneumonia, Kidney disease, Suicide.*"

"*To create this dataset use the same proportions of the means seen below:*"

Year, Avg_death_rate, Avg_deaths:

1999 144 15025
2000 142 15054
2001 140 15090
2002 140 15253
2003 138 15246
2004 132 14901
2005 133 15168
2006 129 14968
2007 126 14930
2008 126 15192
2009 122 14950
2010 121 15108
2011 120 15331
2012 118 15454
2013 118 15760
2014 118 15961
2015 119 16523
2016 119 16708
2017 119 17116

Average (Death Rate, Average Number of Deaths):

Alabama (153, 7672)
Alaska (123, 564)
Arizona (117, 7524)
Arkansas (148, 4725)
California (112, 39089)
Colorado (113, 4973)
Connecticut (111, 4694)
Delaware (128, 1220)
District of Columbia (137, 808)
Florida (116, 28078)
Georgia (138, 11130)
Hawaii (99.2, 1534)
Idaho (121, 1797)
Illinois (126, 16698)
Indiana (138, 9257)
Iowa (120, 4567)
Kansas (126, 3980)
Kentucky (152, 6813)
Louisiana (151, 6744)
Maine (126, 2074)
Maryland (123, 7032)
Massachusetts (114, 8671)
Michigan (132, 14408)
Minnesota (108, 6163)
Mississippi (159, 4747)
Missouri (138, 9117)
Montana (125, 1410)
Nebraska (119, 2455)
Nevada (134, 3123)
New Hampshire (119, 1693)
New Jersey (118, 11469)
New Mexico (123, 2494)
New York (115, 24694)
North Carolina (134, 12479)
North Dakota (116, 967)
Ohio (137, 17781)
Oklahoma (152, 5948)
Oregon (120, 5091)
Pennsylvania (129, 20596)
Rhode Island (120, 1568)
South Carolina (140, 6530)
South Dakota (121, 1168)
Tennessee (149, 9628)
Texas (128, 26483)
United States (126, 401958)
Utah (114, 2261)
Vermont (119, 858)
Virginia (125, 9518)
Washington (118, 7823)
West Virginia (154, 3454)
Wisconsin (120, 7690)
Wyoming (127, 695)

"*Some requirements is to create exactly 200 rows per state, preserve yearly and state level averages approximately, use set.seed() for reproducibility, store number of Deaths and "Age-adjusted Death Rate" as numeric export final dataset to CSV and use tidyverse code,*"

**Generating synthetic dataset:**

```{r}
library(tidyverse)

set.seed(123)

years <- 1999:2017

year_targets <- tribble(
  ~Year, ~Avg_death_rate, ~Avg_deaths,
  1999, 144, 15025,
  2000, 142, 15054,
  2001, 140, 15090,
  2002, 140, 15253,
  2003, 138, 15246,
  2004, 132, 14901,
  2005, 133, 15168,
  2006, 129, 14968,
  2007, 126, 14930,
  2008, 126, 15192,
  2009, 122, 14950,
  2010, 121, 15108,
  2011, 120, 15331,
  2012, 118, 15454,
  2013, 118, 15760,
  2014, 118, 15961,
  2015, 119, 16523,
  2016, 119, 16708,
  2017, 119, 17116
)

state_targets <- tribble(
  ~State, ~state_mean_rate, ~state_mean_deaths,
  "Alabama",153,7672,
  "Alaska",123,564,
  "Arizona",117,7524,
  "Arkansas",148,4725,
  "California",112,39089,
  "Colorado",113,4973,
  "Connecticut",111,4694,
  "Delaware",128,1220,
  "District of Columbia",137,808,
  "Florida",116,28078,
  "Georgia",138,11130,
  "Hawaii",99.2,1534,
  "Idaho",121,1797,
  "Illinois",126,16698,
  "Indiana",138,9257,
  "Iowa",120,4567,
  "Kansas",126,3980,
  "Kentucky",152,6813,
  "Louisiana",151,6744,
  "Maine",126,2074,
  "Maryland",123,7032,
  "Massachusetts",114,8671,
  "Michigan",132,14408,
  "Minnesota",108,6163,
  "Mississippi",159,4747,
  "Missouri",138,9117,
  "Montana",125,1410,
  "Nebraska",119,2455,
  "Nevada",134,3123,
  "New Hampshire",119,1693,
  "New Jersey",118,11469,
  "New Mexico",123,2494,
  "New York",115,24694,
  "North Carolina",134,12479,
  "North Dakota",116,967,
  "Ohio",137,17781,
  "Oklahoma",152,5948,
  "Oregon",120,5091,
  "Pennsylvania",129,20596,
  "Rhode Island",120,1568,
  "South Carolina",140,6530,
  "South Dakota",121,1168,
  "Tennessee",149,9628,
  "Texas",128,26483,
  "United States",126,401958,
  "Utah",114,2261,
  "Vermont",119,858,
  "Virginia",125,9518,
  "Washington",118,7823,
  "West Virginia",154,3454,
  "Wisconsin",120,7690,
  "Wyoming",127,695
)

states <- state_targets$State
##Causes of death
causes_map <- tribble(
  ~`Cause Name`, ~`113 Cause Name`,
  "All causes", "All causes (A00-Y89)",
  "Heart disease", "Diseases of heart (I00-I09,I11,I13,I20-I51)",
  "Cancer", "Malignant neoplasms (C00-C97)",
  "Stroke", "Cerebrovascular diseases (I60-I69)",
  "CLRD", "Chronic lower respiratory diseases (J40-J47)",
  "Unintentional injuries", "Accidents (V01-X59,Y85-Y86)",
  "Alzheimer’s disease", "Alzheimer disease (G30)",
  "Diabetes", "Diabetes mellitus (E10-E14)",
  "Influenza and pneumonia", "Influenza and pneumonia (J09-J18)",
  "Kidney disease", "Nephritis (N00-N07,N17-N19)",
  "Suicide", "Intentional self-harm (X60-X84,Y87.0)"
)

# Give causes relative weights (mean normalized to 1)
cause_scales <- tibble(
  `Cause Name` = causes_map$`Cause Name`,
  weight = c(1.6,1.25,1.2,0.7,0.7,0.9,0.75,0.75,0.65,0.65,0.6)
) %>%
  mutate(scale = weight / mean(weight)) %>%
  select(-weight)

#Build expected state/year means

year_multipliers <- year_targets %>%
  mutate(
    year_mult_deaths = Avg_deaths / mean(Avg_deaths),
    year_mult_rate   = Avg_death_rate / mean(Avg_death_rate)
  ) %>%
  select(Year, year_mult_deaths, year_mult_rate)

expected <- crossing(State = states, Year = years) %>%
  left_join(state_targets, by = "State") %>%
  left_join(year_multipliers, by = "Year") %>%
  mutate(
    expected_deaths = state_mean_deaths * year_mult_deaths,
    expected_rate   = state_mean_rate * year_mult_rate
  ) %>%
  select(State, Year, expected_deaths, expected_rate)

#Generate exactly 200 Rows Per State
n_per_state <- 200

year_probs <- year_targets %>%
  mutate(p = Avg_deaths / sum(Avg_deaths))

raw_data <- tibble(State = states) %>%
  mutate(
    Year = map(State, ~ sample(year_probs$Year, n_per_state, replace = TRUE, prob = year_probs$p)),
    `Cause Name` = map(State, ~ sample(causes_map$`Cause Name`, n_per_state, replace = TRUE))
  ) %>%
  unnest(c(Year, `Cause Name`)) %>%
  left_join(causes_map, by = "Cause Name") %>%
  left_join(cause_scales, by = "Cause Name") %>%
  left_join(expected, by = c("State","Year")) %>%
  mutate(
    mu_deaths = expected_deaths * scale,
    mu_rate   = expected_rate * scale,
    Deaths = rlnorm(n(), log(pmax(mu_deaths,1)), 0.35),
    `Age-adjusted Death Rate` = rlnorm(n(), log(pmax(mu_rate,0.1)), 0.18)
  ) %>%
  select(Year, `113 Cause Name`, `Cause Name`, State, Deaths, `Age-adjusted Death Rate`)

calibrate <- function(df, iters = 5) {

  reduce(
    seq_len(iters),
    .init = df,
    .f = function(current, i){

      # State scaling
      current <- current %>%
        left_join(state_targets, by = "State") %>%
        group_by(State) %>%
        mutate(
          Deaths = Deaths * (state_mean_deaths / mean(Deaths)),
          `Age-adjusted Death Rate` =
            `Age-adjusted Death Rate` *
            (state_mean_rate / mean(`Age-adjusted Death Rate`))
        ) %>%
        ungroup() %>%
        select(-state_mean_rate, -state_mean_deaths)

      # Year scaling
      current <- current %>%
        left_join(year_targets, by = "Year") %>%
        group_by(Year) %>%
        mutate(
          Deaths = Deaths * (Avg_deaths / mean(Deaths)),
          `Age-adjusted Death Rate` =
            `Age-adjusted Death Rate` *
            (Avg_death_rate / mean(`Age-adjusted Death Rate`))
        ) %>%
        ungroup() %>%
        select(-Avg_death_rate, -Avg_deaths)

      current
    }
  )
}

synthetic_data <- raw_data %>%
  calibrate(iters = 6) %>%
  mutate(
    Deaths = as.numeric(round(pmax(Deaths,0))),
    `Age-adjusted Death Rate` = as.numeric(round(`Age-adjusted Death Rate`,1))
  ) %>%
  arrange(State, Year)

write_csv(
  synthetic_data,
  here("cdcdata-exercise/synthetic_mortality_dataset.csv")
)

synthetic_data %>% count(State)    # 200 rows per state
synthetic_data %>% count(State, Year) %>% head()
```

Now, let's inspect the data: 

```{r}
library(readr)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(here)
library(gridExtra)

data <- read_csv(
  here("synthetic_mortality_dataset.csv"),
  locale = locale(encoding = "UTF-8"),
  show_col_types = FALSE
)

colnames(data)
summary(data)

clean_data <- data

```

Making sure there is the same number of reports per state (200): 

```{r}

ggplot(clean_data, aes(x = State)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90))

```

Now I will analyze this dataset using some approaches previously used by Emil, starting with the number of deaths per cause of death over the years:

```{r}

clean_data %>% 
  group_by(`Cause Name`) %>%   
  filter(State == "United States") %>%
  summarise(total_deaths = sum(Deaths)) %>%
  arrange(desc(total_deaths))

P1 <- clean_data %>%
  filter(State == "United States") %>%
  filter(`Cause Name` == "Heart disease") %>%   
  ggplot(aes(x = Year, y = Deaths)) +
  geom_line() +
  theme_classic() +
  labs(title = "Heart Disease")

P2 <- clean_data %>%
  filter(State == "United States") %>%
  filter(`Cause Name` == "Cancer") %>%
  ggplot(aes(x = Year, y = Deaths)) +
  geom_line() +
  theme_classic() +
  labs(title = "Cancer")

P3 <- clean_data %>%
  filter(State == "United States") %>%
  filter(`Cause Name` == "Stroke") %>%
  ggplot(aes(x = Year, y = Deaths)) +
  geom_line() +
  theme_classic() +
  labs(title = "Stroke")

P4 <- clean_data %>%
  filter(State == "United States") %>%
  filter(`Cause Name` == "CLRD") %>%
  ggplot(aes(x = Year, y = Deaths)) +
  geom_line() +
  theme_classic() +
  labs(title = "CLRD")

P5 <- clean_data %>%
  filter(State == "United States") %>%
  filter(`Cause Name` == "Unintentional injuries") %>%
  ggplot(aes(x = Year, y = Deaths)) +
  geom_line() +
  theme_classic() +
  labs(title = "Unintentional Injuries")

P6 <- clean_data %>%
  filter(State == "United States") %>%
  filter(str_detect(`Cause Name`, "Alzheimer")) %>%
  ggplot(aes(x = Year, y = Deaths)) +
  geom_line() +
  theme_classic() +
  labs(title = "Alzheimer's Disease")

P7 <- clean_data %>%
  filter(State == "United States") %>%
  filter(`Cause Name` == "Diabetes") %>%
  ggplot(aes(x = Year, y = Deaths)) +
  geom_line() +
  theme_classic() +
  labs(title = "Diabetes")
    
P8 <- clean_data %>%
  filter(State == "United States") %>%
  filter(`Cause Name` == "Influenza and pneumonia") %>%
  ggplot(aes(x = Year, y = Deaths)) +
  geom_line() +
  theme_classic() +
  labs(title = "Influenza and Pneumonia")

P9 <- clean_data %>%
  filter(State == "United States") %>%
  filter(`Cause Name` == "Kidney disease") %>%
  ggplot(aes(x = Year, y = Deaths)) +
  geom_line() +
  theme_classic() +
  labs(title = "Kidney Disease")

P10 <- clean_data %>%
  filter(State == "United States") %>%
  filter(`Cause Name` == "Suicide") %>%
  ggplot(aes(x = Year, y = Deaths)) +
  geom_line() +
  theme_classic() +
  labs(title = "Suicide")

grid.arrange(P1, P2, P3, P4, P5, P6, P7, P8, P9, P10,
             ncol = 2, nrow = 5)
```


```{r}
#Let's see how they all did when graphed together.
clean_data %>%
  filter(State == "United States") %>%
  filter(`Cause Name` != "All causes") %>%
  ggplot(aes(x = Year,
             y = Deaths,
             group = `Cause Name`,
             color = `Cause Name`)) +
  geom_line() +
  theme_classic()
```

This next piece of analysis examines the number of cancer deaths throughout the years:

```{r}
P2b <- clean_data %>%
  filter(State == "United States") %>%
  filter(`Cause Name` == "Cancer") %>%
  ggplot(aes(x = Year, y = `Age-adjusted Death Rate`)) +
  geom_line() +
  theme_classic()

grid.arrange(P2, P2b, ncol = 2, nrow = 1)
```

**Observations based on real data analyzed by Emil and synthetic data generated:**

- Overall, the data follow a similar pattern over the years in terms of peaks, decreases, and increases in the number of deaths. However, the synthetic dataset appears more variable in its progression over time across all generated plots as we can see in the second figure showing all causes of death death x number of cases over time. 

- In the final plot, which shows the number of deaths caused by Cancer over time, it is possible to observe the same general pattern previously mentioned: an increase in the early 2000s, a decrease around 2005, and another increase in subsequent years. In both the real and simulated datasets, the age-adjusted death rate shows a general reduction over time.

